---
title: "Simulating from mixed-effects general hazard models (MEGH)"
author: '[F. Javier Rubio](https://sites.google.com/site/fjavierrubio67/) and Reza Drikvandi'
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
bibliography: references.bib
---

# MEGH Structure

The MEGH model was proposed by @rubio:2021 in terms of the hazard function:
Let $o_{ij} \in {\mathbb R}_+$, $j=1,\dots,n_i$, $i=1,\dots,r$, be a sample of  times-to-event of interest, and ${\bf x}_{ij}\in{\mathbb R}^p$ be a vector of covariates associated to the $ij$th individual. Let $c_{ij} \in {\mathbb R}_+$ be the right-censoring indicators, and $t_{ij} = \min\{o_{ij},c_{ij}\}$ be the observed survival times. Let $d_{ij} = I(o_{ij} < c_{ij})$ be the vital status indicators (1 - dead, 0 - alive), and $n = \sum_{i = 1}^r n_i$ be the total sample size.
We propose the mixed-effects general hazards (MEGH) model, which is defined through the individual hazard function:
$$
h(t_{ij}\mid {\bf x}_{ij}, {\bf u}_i, \tilde{{\bf u}}_i) = h_0\left(t_{ij} \exp\left\{\tilde{{\bf x}}_{ij}^{\top}\mathbf{\alpha} + \tilde{{\bf z}}_{ij}^{\top}\tilde{{\bf u}}_i \right\}\right)\exp\left\{{\bf x}_{ij}^{\top}\mathbf{\beta} + {\bf z}_{ij}^{\top}{\bf u}_i \right\},	
$$
where $h_0(\cdot)$ is a baseline hazard, $\mathbf{\beta}=(\beta_1,\ldots,\beta_p)^\top \in \mathbb{R}^p$, $\mathbf{\alpha}=(\alpha_1,\ldots,\alpha_q)^\top \in \mathbb{R}^{\tilde{p}}$, 
${{\bf x}}_{ij}\in \mathbb{R}^p$ is a vector of covariates affecting the hazard scale, $\tilde{{\bf x}}_{ij} \in \mathbb{R}^{\tilde{p}}$ is a vector of covariates affecting the time scale (typically, $\tilde{{\bf x}}_{ij} \subset {\bf x}_{ij}$), and $({\bf u}_i,\tilde{{\bf u}}_i) \stackrel{iid}{\sim} G_{q+\tilde{q}}$, where $G_{q+\tilde{q}}$ is a continuous distribution with support on ${\mathbb R}^{q + \tilde{q}}$. 

Suppose that $q=\tilde{q} = 1$ and that $G_{q+\tilde{q}}$ is a bivariate normal distribution with zero mean and covariance matrix $\Sigma$. The following R code shows how to simulate from the MEGH model under these assumptions for specific values of the parameters and a baseline hazard given by the [Power Generalised Weibull](https://rpubs.com/FJRubio/PGW) distribution. In order to avoid simulating design matrices, we will use the real data set `LeukSurv` from the R package `spBayesSurv`.

```{r}
##############################################################################################
# MEGH
##############################################################################################

# Required packages
library(mvtnorm)
#library(devtools)
#install_github("FJRubio67/MEGH")
library(MEGH)
library(spBayesSurv)

#********************************************************
# Design matrices and ID indicators based on real data
#********************************************************

data(LeukSurv)
#View(LeukSurv)
#dim(LeukSurv)

X <- as.matrix(cbind(scale(LeukSurv$age), LeukSurv$sex, scale(LeukSurv$wbc), scale(LeukSurv$tpi) ))
colnames(X) <- cbind("std age", "sex", "wbc", "tpi")
Xt <- as.matrix(cbind(scale(LeukSurv$age)))
n.ind <- dim(X)[1]  # number of individuals
n.clust <- length(unique(LeukSurv$district)) # number of clusters
q <- dim(Xt)[2]
p <- dim(X)[2]
r <- n.clust
ID <- as.vector(LeukSurv$district)


#True values of the parameters
Sigma <- cbind(c(1,0.5),c(0.5,1))
alpha <- 0.96 
beta <- c(1.00,0.08,0.22,0.10) 
theta <- c(0.2,1.50,3.0)
true_parameters <- c(sigma,theta,alpha,beta)
npar <- length(true_parameters)
cens <- 4 # censoring times


#********************************************************
# Simulation
#********************************************************
#*
simulated_survival <- simMEGH(seed=1234, des=X, des_t=Xt, ID=ID, 
                              alpha=alpha, beta=beta, theta=theta, Sigma=Sigma, 
                              cens=rep(cens,n.ind), restr="GENERAL", distr="PGW")  

summary(simulated_survival)

# Histogram of survival times
hist(simulated_survival$times, xlab = "times", ylab ="density", probability = TRUE, main = "")
box()

# Proportion of observed survival times
mean(simulated_survival$status)

```

# MEGH-I and MEGH-II sub-structures

@rubio:2021 proposed two sub-structures of the MEGH structure.

## MEGH-I Structure
The first proposed mixed effects hazard structure is:
$$
h_1(t_{ij}\mid {\bf x}_{ij}, {\bf u}_i) = h_0\left(t_{ij} \exp\left\{\tilde{{\bf x}}_{ij}^{\top}\mathbf{\alpha}\right\}\right)\exp\left\{{\bf x}_{ij}^{\top}\mathbf{\beta} + {\bf z}_{ij}^{\top}{\bf u}_i \right\},	
$$

This mixed hazard structure contains the mixed effects proportional hazards (MEPH, $\mathbf{\alpha} = 0$) model @therneau:2000, while allowing for the inclusion of time-dependent effects through the parameter $\mathbf{\alpha}$.  

The following R code shows how to simulate from the MEGH-I model under these assumptions for specific values of the parameters and a baseline hazard given by the [Power Generalised Weibull](https://rpubs.com/FJRubio/PGW) distribution. In order to avoid simulating design matrices, we will use the real data set `LeukSurv` from the R package `spBayesSurv`.

```{r}
##############################################################################################
# MEGH-I
##############################################################################################

# Required packages
library(mvtnorm)
library(MEGH)
library(spBayesSurv)

#********************************************************
# Design matrices and ID indicators based on real data
#********************************************************

data(LeukSurv)
#View(LeukSurv)
#dim(LeukSurv)

X <- as.matrix(cbind(scale(LeukSurv$age), LeukSurv$sex, scale(LeukSurv$wbc), scale(LeukSurv$tpi) ))
colnames(X) <- cbind("std age", "sex", "wbc", "tpi")
Xt <- as.matrix(cbind(scale(LeukSurv$age)))
n.ind <- dim(X)[1]  # number of individuals
n.clust <- length(unique(LeukSurv$district)) # number of clusters
q <- dim(Xt)[2]
p <- dim(X)[2]
r <- n.clust
ID <- as.vector(LeukSurv$district)


#True values of the parameters
Sigma <- 1
alpha <- 0.96 
beta <- c(1.00,0.08,0.22,0.10) 
theta <- c(0.2,1.50,3.0)
true_parameters <- c(sigma,theta,alpha,beta)
npar <- length(true_parameters)
cens <- 4 # censoring times


#********************************************************
# Simulation
#********************************************************
#*
simulated_survival <- simMEGH(seed=1234, des=X, des_t=Xt, ID=ID, 
                              alpha=alpha, beta=beta, theta=theta, Sigma=Sigma, 
                              cens=rep(cens,n.ind), restr="I", distr="PGW")  

summary(simulated_survival)

# Histogram of survival times
hist(simulated_survival$times, xlab = "times", ylab ="density", probability = TRUE, main = "")
box()

# Proportion of observed survival times
mean(simulated_survival$status)
```


## MEGH-II Structure
The second proposed mixed effects hazard structure is:
$$
h_2(t_{ij}\mid {\bf x}_{ij}, {\bf u}_i) = h_0\left(t_{ij} \exp\left\{\tilde{{\bf x}}_{ij}^{\top}\mathbf{\alpha} + {\bf z}_{ij}^{\top}{\bf u}_i \right\}\right)\exp\left\{{\bf x}_{ij}^{\top}\mathbf{\beta} + {\bf z}_{ij}^{\top}{\bf u}_i \right\},	
$$

This mixed hazard structure generalises the mixed effects accelerated failure time (MEAFT, $\mathbf{\alpha}=\mathbf{\beta}$) structure (see \emph{e.g.}~@rubio:2018b). Moreover, if no random effects are included, then we obtain the proportional hazards (PH) model for $\mathbf{\alpha}=0$, the accelerated failure time (AFT) model for $\mathbf{\alpha}=\mathbf{\beta}$, and the accelerated hazards (AH) model for $\mathbf{\beta}=0$ \citep{chen:2001}. Figure \ref{fig:diag} shows the link between MEGH and some submodels of interest. Note that the MEGH is much richer than that, but this structure illustrates the generality of the proposed mixed structure.

The following R code shows how to simulate from the MEGH-II model under these assumptions for specific values of the parameters and a baseline hazard given by the [Power Generalised Weibull](https://rpubs.com/FJRubio/PGW) distribution. In order to avoid simulating design matrices, we will use the real data set `LeukSurv` from the R package `spBayesSurv`.

```{r}
##############################################################################################
# MEGH-II
##############################################################################################

# Required packages
library(mvtnorm)
library(MEGH)
library(spBayesSurv)

#********************************************************
# Design matrices and ID indicators based on real data
#********************************************************

data(LeukSurv)
#View(LeukSurv)
#dim(LeukSurv)

X <- as.matrix(cbind(scale(LeukSurv$age), LeukSurv$sex, scale(LeukSurv$wbc), scale(LeukSurv$tpi) ))
colnames(X) <- cbind("std age", "sex", "wbc", "tpi")
Xt <- as.matrix(cbind(scale(LeukSurv$age)))
n.ind <- dim(X)[1]  # number of individuals
n.clust <- length(unique(LeukSurv$district)) # number of clusters
q <- dim(Xt)[2]
p <- dim(X)[2]
r <- n.clust
ID <- as.vector(LeukSurv$district)


#True values of the parameters
Sigma <- 1
alpha <- 0.96 
beta <- c(1.00,0.08,0.22,0.10) 
theta <- c(0.2,1.50,3.0)
true_parameters <- c(sigma,theta,alpha,beta)
npar <- length(true_parameters)
cens <- 4 # censoring times


#********************************************************
# Simulation
#********************************************************
#*
simulated_survival <- simMEGH(seed=1234, des=X, des_t=Xt, ID=ID, 
                              alpha=alpha, beta=beta, theta=theta, Sigma=Sigma, 
                              cens=rep(cens,n.ind), restr="II", distr="PGW")  

summary(simulated_survival)

# Histogram of survival times
hist(simulated_survival$times, xlab = "times", ylab ="density", probability = TRUE, main = "")
box()

# Proportion of observed survival times
mean(simulated_survival$status)


```

# References
